---
title: "Day 1: Basic R programming"
author: "Markus Loecher, Berlin School of Economics and Law"
output:
   html_document:
    #variant: markdown_github
    toc: true
    number_sections: true
    self_contained: no
    toc_depth: 2
    toc_float: true
    fig_caption: true

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 6, fig.height = 4)
options(digits=3)
```


# data types, vectors, matrices, strings, factors

We need to be very clear on the different **data types** in R!

**numbers**:

```{r}
x = 2
y = 3
x+y
```


**vectors**:

```{r}
x = c(1,2,3)
y = 10:12
#or 
y = seq(10,12,by=1)
#operations on x
x[2]
x[-2]
2*x
x^2
#built-in functions
mean(x)
sum(x)
#vectorized operations
x+y
#bare bones graphics
plot(x,y); grid()
barplot(x); grid()
```

**matrices**:

```{r}
x = matrix(1:9, ncol=3)
colnames(x) = c("var1","var2","var3")
x
dim(x)
x[2,3]
x[2,"var1"]
colMeans(x)
```

**data frames**:

```{r}
x = data.frame(a = 1:3, l = LETTERS[1:3], r = runif(3))
x
x$l

```

**lists**:

```{r}
y = list(x = matrix(1:9, ncol=3), l = LETTERS[1:3], r = NA)
y

```


**strings**:

```{r}
x = "chevrolet"
class(x)
```

### factors in R


Try class("chevrolet")

In the Auto set take a look at the data class of the name column:

```{r,echo=TRUE}
library(ISLR)
data(Auto)

class(Auto$name)
```

Try

as.numeric("chevrolet")

Try 

as.numeric(head(Auto$name))


Look at the help file for read.csv and look for the argument *stringsAsFactors*

You can force a variable to be a factor:

```{r,echo=TRUE}
Auto$cylinders = as.factor(Auto$cylinders)
```

#### Understanding factors is extremely important for modeling ! (=dummy variables)


**dates/times**:

```{r}
x = Sys.Date()
#or
as.Date("2017-03-20")
y = Sys.time()
x
y
class(x)
class(y)

as.numeric(x)
as.numeric(y)

#or
unclass(x)
unclass(y)
```


------------------------------

Let us look at some real data now. 
We load in the *S&P Stock Market Data* dataset from the *ISLR* library:

```{r}
library(ISLR)
data(Smarket)

head(Smarket)

mean(Smarket$Lag1)
```

* **Year** The year that the observation was recorded

* **Lagi**: Percentage return for i days previous

* **Volume**: Volume of shares traded (number of daily shares traded in billions)

* **Today**: Percentage return for today

* **Direction**: A factor with levels Down and Up indicating whether the market had a positive or negative return on a given day


This is a data frame with **mixed data types**:


```{r}
dim(Smarket)
class(Smarket)
class(Smarket$Lag1)

```

Let us add a proper date variable:
```{r}
Smarket$day = as.Date("2001-01-01") + 1:1250
```

---------------------------------

# control structures (for, if else, while)


In a non-vectorized language the above command would have necessitated a loop!
Since they are useful on their own in R, we will reproduce the *day* column:

```{r}
#initialize
Smarket$day = NA

for (i in 1:1250) Smarket$day[i] = as.Date("2001-01-01") + i

#or with multiple statements
for (i in 1:1250) {
  Smarket$day[i] = as.Date("2001-01-01") + i
  if (Smarket$day[i] == "2001-07-04") cat("The return on July 4th was ", Smarket$Today[i], "\n")
}

```

### Wait a second, shouldn't we exclude weekends and holidays?

Can we download those from Yahoo finance?

```{r, message=FALSE}
library(tseries)
 # sp500 = get.hist.quote(instrument = "^gspc", "2001-01-02", "2005-12-30",
 #                 quote = c("Open", "Volume"))
 #  save(sp500, file="sp500.rda")
load("sp500.rda")
n=nrow(sp500)
#can we compute percent returns?
sp500$Today = round(100*diff(sp500[,"Open"])/sp500[-n,"Open"],3)
colnames(sp500)[1]= "Open"
```

### Adding lagged variables

```{r}
sp500$Lag1 = lag(sp500$Today,-1)
sp500$Lag2 = lag(sp500$Today,-2)
sp500$Lag3 = lag(sp500$Today,-3)
sp500$Lag4 = lag(sp500$Today,-4)
sp500$Lag5 = lag(sp500$Today,-5)
```


### Task: Use a for loop to add lagged variables

-----------------------------------

# functions, R markdown

So far we have written code that is executed either line by line or as a whole but lacks any "modular" or "reusable" attribute. 
If you want to build anything more complex it is essential to use "building blocks" that work on their own and can be connected to or used by other modules.

You have been using the many functions that are part of R already, such as *hist()*, *boxplot()*, etc. 
Clearly, each of these function consists of many lines of codes that you are actually not all that interested in. 
R being open source, you can just type the name of many functions and you will see the source code, try e.g. 
```{r}
sd
```

Functions consist of

1. a name/title
2. arguments
3. body
4. return values

For example, if you were to write your own summing function, it could look like this 

```{r}
mysum = function(a=4,b=3){
  c = a+b
  print(c)
}

mysum(10,20)

c = mysum(b=2,a=3)
```


Add a few checks and a return value:

```{r}
mysum = function(a=4,b=3){
  #stopifnot(is.numeric(a))
  if (!is.numeric(a)){
    print("a has to be numeric !!!!!! you idiot")
    return()
  }
  c = a+b
  return(c)
}

mysum("Karl",20)

```


Comments:

1. default values
2. named arguments
3. invisible/optional returns
4. local variables


-------------------------


# basic plotting 

### scatter plots

```{r}

plot(mpg ~ weight, data = Auto,col = rgb(0,0,1,0.5), pch=20);grid()

##ggplot version
library(ggplot2)
p <- ggplot(Auto, aes(weight, mpg))
p + geom_point(color="firebrick", alpha=0.5)
```


### histograms

Back to the Smarket data: can we take a look at the distribution of the data?

```{r, fig.width=10}
par(mfrow=c(1,2))
hist(Smarket$Lag1)
hist(Smarket$Volume)

```

### boxplots

```{r,echo=TRUE}
library(ISLR)
data(Auto)

boxplot(mpg ~ cylinders, data=Auto, xlab="cylinders", ylab="mpg");grid()
#notched version - for a different data set
boxplot(mpg ~ cylinders, data=subset(Auto, cylinders %in% c(4,6,8)), xlab="cylinders", ylab="mpg", notch=TRUE);grid()
##ggplot version
library(ggplot2)
p <- ggplot(Auto, aes(factor(cylinders), mpg))
p + geom_boxplot(fill="darkseagreen4")

```

### violin plots

```{r}

p + geom_violin() + geom_jitter(alpha=0.5, width=0.3, aes(color=cylinders))
```


### time series plots

```{r}
plot(sp500[,1:2])
```

### interactive paning and zooming:

```{r}
library(dygraphs)
dy=dygraph(sp500[,"Open"])
dyRangeSelector(dy)
```


--------------------------

# computing with probabilities and distributions



Many statistics text books use a coin flipping example as a gentle introduction to the binomial distribution. We would like to achieve the following tasks:

1. sample from a binary vector, e.g. c(0,1) N times, possibly with $p \neq 0.5$
2. compute the mean or sum
3. repeat steps 1 and 2 MANY times, e.g. M=1000
4. possibly plot a histogram and compute tail probabilities
5. somehow pass some of these measures to the user

The R code would look a bit like this:
```{r}
N=40;M=1000;p=0.5;Head2Tail=0.7
  set.seed(123)
  #initialize parameters:
  coin = c(0,1)#values to sample from
  xm = vector();#sample mean or sum
  #loop
  for (i in 1:M){
    x = sample(coin, N, rep=TRUE, prob  = c(1-p,p))
    xm[i] = mean(x)
  }
  hist(xm)
  #get the 5% and 95% quantiles:
  q = quantile(xm,c(0.05,0.95))
  #add vertical lines:
  abline(v=q,col=2)
  pObs = sum(xm >= Head2Tail) + sum(xm <= 1-Head2Tail)
  
  pObs = sum(xm >= Head2Tail | xm <= 1-Head2Tail)
  #pObs = sum(xm >= Head2Tail & xm <= 1-Head2Tail)
  
```

### Task: Change the code above into a function 

## Binomial and Normal Distributions

All standard distributions such as normal, binomial, possion, gamma etc. are built into R.
Each distribution comes in 4 version:

1. *Density*:  **d**norm(), **d**binom(), **d**pois()
2. *distribution*: **p**norm(), **p**binom(), **p**pois()
3. *quantile*: **q**norm(), **q**binom(), **q**pois()
4. *random generation*:   **r**norm(), **r**binom(), **r**pois()

For example:

```{r}

qnorm(0.975)
pnorm(-1.96)
hist(rnorm(500))


# Compute P(45 < X < 55) for X Binomial(100,0.5)
sum(dbinom(46:54, 100, 0.5))
# Compute P( X < 55) for X Binomial(100,0.5)
pbinom(54, 100, 0.5)

```


#### Task: Create an Rmd file that contains text, code and figures answering the following questions:

1. Random Walker: Imagine you take a succession of 40 random left/right steps of each 1m length. What is the probability that you end up at least 10 meter away from the center? 

2. Repeat the exercise with 400 steps and 100 meters.

3. Size matters: insurance company A insures 100 cars, company B 400 cars. The probability of a car being stolen is 10%. Compute the probabilities that more than 15% of the respective fleets are stolen.

4. Faced with a mutliple choice test containing 20 question with 4 choices each you decide in desparation to just guess all answers. What is the probability that you will pass, i.e. get at least 10 correct answers?


5. At a certain intersection, the light for eastbound traffic is red for $15$ seconds, yellow for $5$ seconds, and green for $30$ seconds.  Find the probability that out of the next twenty eastbound cars that arrive randomly at the light, fewer than three will be stopped by a red light.


-----------------------------------

# file I/O

* `scan()`
* `read.csv()`, `read.table()` 
* `load()`
* "Import Dataset"

-----------------------------------

#### Task: swirl exercises





